---
title: "project1"
output: pdf_document
---

Data project 1, Due Feb 28.



Instructions from part1-instructions-1.pdf

* each answer in a new page
* You must indicate on Gradescope which questions are on which pages.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## install pacman to help manage installation and loading of additional libraries
    
    Optional part, and it doesn't seems to play well with knit :-\
    https://www.rdocumentation.org/packages/pacman/versions/0.5.1
    pacman provides wrapper function like p_load() to install package if needed, then load the library 
    
```{r install_pacman}
    #install.packages( "pacman" )
    #library(pacman) 
    #p_load( readxl )
```


## Loading Ghana data 
data access: https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.dr130 

the docx explaining what each cols are:

id      participant's ID
age     age
sex     sex
geo     urban or rural resident
pai     plasminogen activator inhibitor-1
tpa     tissue plasminogen activator
bmi     body mass index
sbp     systolic blood pressure
dbp     diastolic blood pressure
chol    total cholesterol
hdl     high-density lipoprotein cholesterol
trig    triglycerides
ldl     low-density lipoprotein cholesterol
tc_hdl  total cholesterol-high-density lipoprotein cholesterol ratio
gluc    fasting glucose
smo     1= smoker, 0=non-smoker
sch     1= some formal education, 0 = none
sch>ss  1= schooling being secondary school, 0= none

```{r ghana}

library(readxl)
library(dplyr)


ghana = read_xlsx("ghana/GHANA_PLoS_upload.xlsx")
summary(ghana)
str(ghana)
ghana_sample = ghana %>% sample_frac(0.1)
#View(ghana_sample)
```

## Including Plots

You can also embed plots, for example:
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


```{r pressure, echo=FALSE}
##plot(ghana)  ## this takes like 3-5 minutes... 
plot(ghana_sample)
```

