---
title: "project1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## install pacman to help manage installation and loading of additional libraries
    
    Optional part, and it doesn't seems to play well with knit :-\
    https://www.rdocumentation.org/packages/pacman/versions/0.5.1
    pacman provides wrapper function like p_load() to install package if needed, then load the library 
    
```{r install_pacman}
    #install.packages( "pacman" )
    #library(pacman) 
    #p_load( readxl )
```


## Loading Ghana data 
data access: https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.dr130 

there is a docx explaining what each cols are


```{r ghana}

library(readxl)
library(dplyr)


ghana = read_xlsx("ghana/GHANA_PLoS_upload.xlsx")
summary(ghana)
str(ghana)
ghana_sample = ghana %>% sample_frac(0.1)
#View(ghana_sample)
```

## Including Plots

You can also embed plots, for example:
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


```{r pressure, echo=FALSE}
##plot(ghana)  ## this takes like 3-5 minutes... 
plot(ghana_sample)
```

